External priors for the next generation of CMB experiments.

### 2016-02-09 Notes/CMBS4_priors_paper.tex
typos fixed

Small scale structure formation is suppressed if fast-moving neutrinos comprise a significant part of the matter budget. 
Indeed, below the current free-streaming scale, the matter power spectrum is suppressed in the presence of
massive neutrinos by a factor $\Delta P/P\simeq -8f_{\nu}$ , where $f_{\nu} = \Omega_{\nu} / \Omega_{m}$ is the contribution of neutrinos to the total matter density.
This effect is probed by CMB lensing and, since the active neutrino number densities are known in the standard model, can be directly transformed into a constraint on the sum of the neutrino masses $\sum m_\nu$. This is particularly exciting because we now know that neutrinos are massive, with a lower limit $\sum m_\nu>50$ meV that emerges from oscillation experiments (see \cite{bellini:2013} and references therein).

%==========
%\begin{figure}[htbp]
%\begin{center}
%\includegraphics{prior_omnuh2.pdf}
%\caption{
%Constraints on the sum of the neutrino masses as a function of priors on cosmological parameters. 
%Each prior is expressed in units of the error that would be obtained internally from the CMB S4 experiment (+Planck+current BAO); e.g., $\sigma(H_0)_{\rm S4 + Planck Pol + BAO15}=0.6\%$, so the value of unity on the x-axis corresponds to imposing an external prior on $H_0$ of 0.6\%.
%The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI. The bottom-most curve shows the impact of imposing the given prior on {\it all} the other parameters simultaneously.
%$H_{0}$ and $\Omega_{c}h^{2}$ are shown to be the most important external priors to improve neutrino mass constraints. We chose a fiducial value $\sum m_{\nu}=85 $ meV.}
%\label{fig:prior_omeganuh2}
%\end{center}
%\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics{{prior_omnuh2_hubble_lmin=4_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
\caption{Constraints on the sum of the neutrino masses (absolute on the left and and relative to its value on the right) as a function of priors on the Hubble parameter, H$_{0}$. The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
\label{fig:prior_omeganuh2-h}
\end{center}
\end{figure}
%==========

\begin{figure}[htbp]
\begin{center}
\includegraphics{{prior_omnuh2_omch2_lmin=4_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
\caption{Constraints on the sum of the neutrino masses (absolute on the left and and relative to its value on the right) as a function of priors on $\Omega_{c} h^{2}$. The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
\label{fig:prior_omeganuh2-omch2}
\end{center}
\end{figure}
%==========

@@

\reffig{prior_omeganuh2-h} and \reffig{prior_omeganuh2-omch2} show the projected errors on $\sum m_\nu$ from CMB-S4 (+Planck+current BAO) as a function of external priors. 
Without any external priors (far right in figure), upcoming CMB experiments are poised to obtain 1-sigma limits on $\sum m_\nu$ of order 25 meV, corresponding to a 2-sigma detection even in the worst case scenario. Adding in external priors helps this significantly though, as discussed, for example, in \cite{di-valentino:2015,2013arXiv1309.5383A,pan:2015a,allison:2015}. 
There, the added prior was ``DESI BAO'', representing a measurement of distances as a function of redshift from the baryon acoustic oscillation feature probed by the planned experiment DESI \cite{levi:2013}. This is represented by the blue band in our figures. In addition to showing the implications of introducing this fixed external constraint, \reffig{prior_omeganuh2-h} and \reffig{prior_omeganuh2-omch2} generalize the extent to which external information will improve the neutrino mass constraints. For example, in \reffig{prior_omeganuh2-h} the red curve shows the improvement as additional information on the Hubble constant is applied. Note that the projected, marginalized constraint on $H_0$ from just CMB-S4+Planck+BAO15 is $0.6\%$ (and similarly the values for the other parameters are given below the figure). So imposing {\it external} information with this value of $\sigma_{H_0}/H_0=0.6\%$ would improve the neutrino mass constraint to 17 meV. Similarly, in \reffig{prior_omeganuh2-omch2} introducing an external constraint on the dark matter density equal to its internal constraint of 0.4\% improves the projected neutrino mass constraint to 20 meV. It is not surprising that this is close to the constraint obtained with DESI because the BAO distances in this model depends most sensitively on $\Omega_ch^2$ and $H_{0}$.

The physical reason underlying this improvement is simple: lensing constrains $f_\nu$, the ratio of the neutrino density to the matter density. In order to extract the neutrino density (and therefore the neutrino mass), we need to know the matter density. A prior on $H_{0}$ helps for a related, but subtle reason. 
The position of the CMB peaks, well constrained by measurement, in a flat Universe depends on the combination $\Omega_ch^{2.93}$ \cite{planck-collaboration:2014} 
. Therefore, an external prior on $H_{0}$ will improve the internal reconstruction of $\Omega_ch^2$, leading back to tighter neutrino mass constraints.
As found in \cite{2013arXiv1309.5383A,pan:2015a,allison:2015}, the projected error falls below 20 meV with these priors. 
The left-hand side of \reffig{prior_omeganuh2-h} and \reffig{prior_omeganuh2-omch2} shows that even tighter external constraints on either $H_0$ or $\Omega_ch^2$ would significantly improve the neutrino mass constraints.

@@

It may be difficult to get the very low-$l$ modes from the ground with CMB-S4, so in Fig.\ref{fig:prior_omeganuh2-50-as}-\ref{fig:prior_omeganuh2-50-ns} we limit the modes to $\ell>\ell_{\rm min}=50$. 
Together with the importance of $H_{0}$ and $\Omega_ch^2$ previously observed, in this case external priors on $A_{s}$ and $\tau$ are found to be helpful (see \reffig{prior_omeganuh2-50-ns} and \reffig{prior_omeganuh2-50-as}).

Indeed, as recently pointed out by \cite{allison:2015}, a degeneracy exists between $\sum m_\nu$ and $A_{s}$ and, as a consequence, between $\sum m_\nu$ and $\tau$. This can be understood as follows.
The CMB lensing reconstruction is noisy at scales larger than the neutrino free streaming scale (see Fig. 5 of \cite{2013arXiv1309.5383A}). For this reason it is hard to measure the unsuppressed CMB lensing power spectra that is then compared to the small scale suppressed one to constrain the values of $f_{\nu}$. The unsuppressed amplitude of the lensing spectrum is actually better constrained by the primordial amplitude $A_{s}$, which is probed by the primordial CMB. Unfortunately, the CMB spectra are sensitive to $A_{s}\exp^{-2\tau}$, so 
a measurement of the optical depth is required to infer $A_{s}$ and therefore to tighten the constraints on $\sum m_\nu$. If CMB S4 will not be designed to measure scales $\ell<50$ to avoid difficulties in dealing with atmospheric noise, an external prior on the optical depth from future 21 cm surveys \cite{liu:2015} or CMB satellite experiments like LiteBIRD \cite{matsumura:2014}, PIXIE\cite{kogut:2011}, and COrE/PRISM \cite{the-core-collaboration:2011,prism-collaboration:2013}, balloon-borne experiments like SPIDER \cite{filippini:2010}, EBEX \cite{reichborn-kjennerud:2010} or eventually ground experiments like CLASS \cite{essinger-hileman:2014}, will be needed to fully exploit the CMB constraining power.

@@

%==========
%\begin{figure}[htbp]
%\begin{center}
%\includegraphics{prior_massless_neutrinos.pdf}
%\caption{Projected 1-sigma error on $N_{\rm eff}$ as a function of external priors. The parameters that would be most useful to constrain externally for the purposes of determining $N_{\rm eff}$ are $n_s$ and $\Omega_bh^2$.}
%\label{fig:prior_massless_neutrinos}
%\end{center}
%\end{figure}
%==========

@@

%==========
%
%\begin{figure}[htbp]
%\begin{center}
%\includegraphics{prior_w.pdf}
%\caption{Constraints on the dark energy equation of state $w$ as a function of external priors on cosmological parameters. 
%An external prior on $H_{0}$ will be crucial to improve the constraint.} 
%\label{fig:prior_w}
%\end{center}
%\end{figure}
%==========

>>>>

Small scale structure formation is suppressed if fast-moving neutrinos comprise a significant part of the matter budget. 
Indeed, below the free-streaming scale, the matter power spectrum is suppressed in the presence of
massive neutrinos by a factor $\Delta P/P\simeq -8f_{\nu}$ , where $f_{\nu} = \Omega_{\nu} / \Omega_{m}$ is the contribution of neutrinos to the total matter density.
This effect is probed by CMB lensing and, since the active neutrino number densities are known in the standard model, can be directly transformed into a constraint on the sum of the neutrino masses $\sum m_\nu$. This is particularly exciting because we now know that neutrinos are massive, with a lower limit $\sum m_\nu>50$ meV that emerges from oscillation experiments (see \cite{bellini:2013} and references therein).

%==========
\begin{figure}[htbp]
\begin{center}
\includegraphics{{prior_omnuh2_hubble_lmin=4_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
\caption{Constraints on the sum of the neutrino masses (absolute on the left and and relative to its value on the right) as a function of priors on the Hubble parameter, H$_{0}$. The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
\label{fig:prior_omeganuh2-h}
\end{center}
\end{figure}
%==========

%==========
\begin{figure}[htbp]
\begin{center}
\includegraphics{{prior_omnuh2_omch2_lmin=4_lmax=4499_ndet=1000000_fsky=0.75}.pdf}
\caption{Constraints on the sum of the neutrino masses (absolute on the left and and relative to its value on the right) as a function of priors on $\Omega_{c} h^{2}$. The horizontal (blue) line correspond to CMB S4 experiment + Planck Pol + DESI.}
\label{fig:prior_omeganuh2-omch2}
\end{center}
\end{figure}
%==========

@@

\reffig{prior_omeganuh2-h} and \reffig{prior_omeganuh2-omch2} show the projected errors on $\sum m_\nu$ from CMB-S4 (+Planck+current BAO) as a function of external priors on $H_{0}$ and $\Omega_{c}h^{2}$. 
Without any external priors (far right in figure), upcoming CMB experiments are poised to obtain 1-sigma limits on $\sum m_\nu$ of order 25 meV, corresponding to a 2-sigma detection even in the worst case scenario. Adding in external priors helps this significantly though, as discussed, for example, in \cite{di-valentino:2015,2013arXiv1309.5383A,pan:2015a,allison:2015}. 
There, the added prior was ``DESI BAO'', representing a measurement of distances as a function of redshift from the baryon acoustic oscillation feature probed by the planned experiment DESI \cite{levi:2013}. This is represented by the blue band in our figures. In addition to showing the implications of introducing this fixed external constraint, \reffig{prior_omeganuh2-h} and \reffig{prior_omeganuh2-omch2} generalize the extent to which external information will improve the neutrino mass constraints. For example, in \reffig{prior_omeganuh2-h} the red curve shows the improvement as additional information on the Hubble constant is applied. We can see, for example that, imposing {\it external} information with at the level of $\sigma_{H_0}/H_0=0.6\%$ would improve the neutrino mass constraint to 17 meV. Similarly, in \reffig{prior_omeganuh2-omch2} adding to CMB S4 + BAO15 an external constraint on the dark matter density equal to its internal constraint of $0.4\%$ improves the projected neutrino mass constraint to 20 meV. It is not surprising that this is close to the constraint obtained with DESI because the BAO distances in this model depends most sensitively on $\Omega_ch^2$ and $H_{0}$.

The physical reason underlying this improvement is simple: lensing constrains $f_\nu$, the ratio of the neutrino density to the matter density. In order to extract the neutrino density (and therefore the neutrino mass), we need to know the matter density. A prior on $H_{0}$ helps for a related, but subtle reason. 
The position of the CMB peaks, well constrained by measurement, in a flat Universe depends on the combination $\Omega_ch^{2.93}$ \cite{planck-collaboration:2014}.
Therefore, an external prior on $H_{0}$ will improve the internal reconstruction of $\Omega_ch^2$, leading back to tighter neutrino mass constraints.
As found in \cite{2013arXiv1309.5383A,pan:2015a,allison:2015}, the projected error falls below 20 meV with these priors. 
The left-hand side of \reffig{prior_omeganuh2-h} and \reffig{prior_omeganuh2-omch2} shows that even tighter external constraints on either $H_0$ or $\Omega_ch^2$ would significantly improve the neutrino mass constraints.

@@

Because of the presence of foregrounds, it may be difficult to get the very low-$\ell$ modes from the ground with CMB-S4, so in Fig.\ref{fig:prior_omeganuh2-50-as}-\ref{fig:prior_omeganuh2-50-tau} we limit the CMB S4 modes to $\ell>\ell_{\rm min}=50$. For the large scale modes $\ell<\ell_{\rm min}=50$ we use Planck at the level of accuracy forecasted by the Planck Blue Book.  
Together with the importance of $H_{0}$ and $\Omega_ch^2$ previously observed, in this case external priors on $A_{s}$ and $\tau$ are found to be helpful (see \reffig{prior_omeganuh2-50-ns} and \reffig{prior_omeganuh2-50-as}).

Indeed, as recently pointed out by \cite{allison:2015}, a degeneracy exists between $\sum m_\nu$ and $A_{s}$ and, as a consequence, between $\sum m_\nu$ and $\tau$. This can be understood as follows.
The CMB lensing reconstruction is noisy at scales larger than the neutrino free streaming scale (see Fig. 5 of \cite{2013arXiv1309.5383A}). For this reason it is hard to measure the unsuppressed CMB lensing power spectra that is then compared to the small scale suppressed one to constrain the values of $f_{\nu}$. The unsuppressed amplitude of the lensing spectrum is actually better constrained by the primordial amplitude $A_{s}$, which is probed by the primordial CMB. Unfortunately, the CMB spectra are sensitive to $A_{s}\exp^{-2\tau}$, so a measurement of the optical depth is required to infer $A_{s}$ and therefore to tighten the constraints on $\sum m_\nu$. If CMB S4 will not be designed to measure scales $\ell<50$ to avoid difficulties in dealing with atmospheric emission, an external prior on the optical depth will be crucial. This can be obtained from future 21 cm surveys \cite{liu:2015} or other CMB experiments, satellite like LiteBIRD \cite{matsumura:2014}, PIXIE\cite{kogut:2011}, and COrE/PRISM \cite{the-core-collaboration:2011,prism-collaboration:2013}, balloon-borne experiments like SPIDER \cite{filippini:2010}, EBEX \cite{reichborn-kjennerud:2010} or eventually ground experiments like CLASS \cite{essinger-hileman:2014}.

@@

@@
